---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This is an example notebook for how to analyze an individual dataset or a group of datasets.

See the `README.md` for more information about how everything is structured.

Make a copy of this notebook and label it with:  
`ds_analysis_<grp_id>_<ds_id>.Rmd`   
So this would be labeled as "ds_analysis_grp1_test_GSE22047.Rmd"

It is located in:   
`code/`  
  `ds_analysis/`  

Then in your local data directory, make a directory for this group or dataset:  
`data/`  
  `group_data/`  
    `<grp_id>/`  
  `individual_data/`   
    `<ds_id>/` - in this case this the study accession


### BEFORE ANALYSIS
0. Fill in the keep/dataset comparison in `tissue_info` or `treated_cells_info`, and update the `ds_list` spreadsheet. You can do this as you look at the datasets individually, this should just help you to figure out what datasets to use. 

Read the description of the study again, you may also want to look it up and see if there's other info available about the study.
We want to be comparing the same sort of effect - in most cases, smokers and non-smokers - in the same tissue. If the tissue doesn't appear accurate, or there's something else going on, note it and move to the next dataset.

If there are less than 10 males or 10 females, write "small" and skip it for now. 

If two studies appear the same or related, we need to be careful about this and we will note it.

For an example of this, go to the `tissue_info` spreadsheet and look at the "small airway epithelium" dataset. I added the "comparison" and "in group" columns. Also add a "downloaded" column for when it is downloaded.
The comparison should say one of the following if we keep it ("in group" column):
* "current vs never"
* "current vs former"
* "current vs former vs never"
* "former vs never"
and then be followed with additional conditions with a ";".
Alternately if it doesn't fit in these categories, you can just describe
e.g. "current vs former; COPD" -OR- if we're not keeping it b/c there are no healthy smokers: "tumor samples". We may come back to these datasets, the goal is just to group for meta-analysis.

Eventually, we add all of the groups and their descriptions to the `ds_list` file and assign a group id. One study can belong to multiple groups.

### METADATA LEVEL
```{r}
require('GEOquery')
require('tidyverse')

# devtools::install_github("tidyverse/googlesheets4") # <-- make sure to use this instead of just install for this package
devtools::install_github("tidyverse/googlesheets4")
require('googlesheets4') 
devtools::install_github("tidyverse/googledrive")
require('googledrive')


#sheets_auth() # you need to run only the first time, this will setup a browser prompt, and then you can read the sheet

# if you have problems, download this spreadsheet locally and use from there
sample_sex_lab <- read_sheet("https://docs.google.com/spreadsheets/d/1lEUVsyXLDcyUQB7mXk0B-LtsWkBn66gKaIV74OFAGVs/edit#gid=463642465")
# sample_sex_lab <- read.csv("sample_sl.csv")
```

1. Download the dataset
```{r}
grp_id <- "grp1_test"
list.ids <- c("GSE17913",  "GSE43079")
study_name <- list.ids[[1]]
# step 1: try GEOQuery for the "GSE" prefix ones
gse.obj1 <- getGEO(list.ids[[1]]) 


# step 2: list ones not in available, we can grab from refine-bio
# the problem is that this is a little slow. You can also download the "E-" ones from ArrayExpress.
# I will send you a file with all the metadata for these... :(
# just make a note in the `tissue_info` file under this tissue that it could not be downloaded yet
```



2. Identify treatment and control data
```{r}
# note: one obj may have mult platforms
# grab phenotype data
ls(gse.obj1)
p1_gpl570 <- pData(gse.obj1$GSE17913_series_matrix.txt.gz)
p1_gpl6804 <- pData(gse.obj1$`GSE22047-GPL6804_series_matrix.txt.gz`)
#p2_gpl570 <- pData(gse.obj2$`GSE43079-GPL570_series_matrix.txt.gz`)
#p2_gpl16419 <- pData(gse.obj2$`GSE43079-GPL16419_series_matrix.txt.gz`)
```



### FOR EACH STUDY/PLATFORM

You will have to identify which columns are present and describe the smoker status and other covariates 
```{r}
colnames(p1_gpl570) 
p1_gpl570 %>% View() # views phenotypic data
# characteristics_ch1.x often have data too -- here they are a mess tho and contain diff fields, so be careful
p1_gpl570 %>% # grabs relevant columns
  select(geo_accession, characteristics_ch1.1, characteristics_ch1.2, characteristics_ch1.3) %>% 
  View()


# always want `geo_accession` <-- this is the sample ID
# grab the columns that seem relevant!

# then make the names friendlier (no spaces!)
smok_vars <- p1_gpl570 %>% 
  select(geo_accession, "copd status:ch1", "smoking status:ch1") %>%
  rename(copd="copd status:ch1", smoker="smoking status:ch1")

smok_vars %>% View()

# what is missing?
# what are the values?
table(smok_vars$copd)
table(is.na(smok_vars$copd)) # looks like absence means no? this is a bad way to encode data so we're going to do something else with it

table(smok_vars$smoker)  # multiple labels mean the same thing "NS", "nonsmoker" (poor metadata), we will fix
table(is.na(smok_vars$smoker)) # all present, yay

# clean up the data within the columns
smok_vars2 <- smok_vars %>%
  mutate(smoker=case_when(
    smoker %in% c("NS", "nonsmoker") ~ "nonsmoker",
    smoker %in% c("S", "smoker") ~ "smoker"
  )) %>%
  mutate(copd=ifelse(is.na(copd), "no", copd))

# what is the breakdown?
table(smok_vars2$copd, smok_vars2$smoker)

# now for this analysis, we want health nonsmoker vs smoker
smok_healthy <- smok_vars2 %>% 
  filter(copd == "no") %>% 
  select(-copd)     

# what other covariates are present?
# -- many of these are NAs!
# -- are there sex labels somewhere? these should match "text_sex", but sometimes they do not
#summary(p1_gpl570)
covariate <- p1_gpl570 %>% 
  select(geo_accession, "serum 25-oh-d:ch1", "ethnicity:ch1", "department of genetic medicine id:ch1" ) %>%
  rename(d_vit="serum 25-oh-d:ch1", eth="ethnicity:ch1", id=`department of genetic medicine id:ch1`)

# it's possible that there are multiple samples for the same person
# look for this! -- here it isn't the case, but it's something to be careful abt
covariate %>%
  group_by(id) %>%
  count() %>%
  arrange(desc(n))

table(covariate$d_vit)
table(is.na(covariate$d_vit)) # missing 205, this doesnt seem useful, but worth looking at to try to figure out what is going on?
# it's possible someone is present only once...
table(covariate$eth) 
table(is.na(covariate$eth)) # missing 156 samples

```

3. Get the number m and number f in each category, as well as the IDs for these
```{r}
study_sex_lab <- sample_sex_lab %>% 
  filter(study_acc==study_name) %>% 
  select(-study_acc)

smok_w_sex_lab <- smok_healthy %>%
  left_join(study_sex_lab, by=c("geo_accession"="sample_acc"))

# how much is missing sex labels?
table(is.na(smok_w_sex_lab$expr_sex)) # 28 are missing, we still have a fair number
smok_w_sex_lab2 <- smok_w_sex_lab %>% filter(!is.na(expr_sex))

# get the sex breakdown
table(smok_w_sex_lab2[,c("expr_sex", "smoker")])

# if there is less than 5 in a category, make note of it on the `tissue_info` spreadsheet in a column, write out this info but we may need to discard for expression analysis
```

```{r}

# double check the distribution of pred scores, we want the two groups to be well separated!
ggplot(smok_w_sex_lab2, aes(x=pred, y=expr_sex))+
  geom_point(aes(col=expr_sex), position=position_jitter(), alpha=0.5) + 
  xlab("P(male)")+
  ylab("")
# we do not want a ton in the middle. this one looks good


# compare to the text labels if they are present
# - switch "expr_sex" to "text_sex" to visualize

# remove any that are mislabeled
smok_w_sex_lab3 <- smok_w_sex_lab2 %>% 
  filter(as.character(text_sex)==as.character(expr_sex) | text_sex == "unknown")
# (here none are removed)
```

4. Write out the breakdown in two places 
```{r}
# another way to get breakdown, we will save this

require('tidyverse')
require('tidyr')

counts_by_grp <- smok_w_sex_lab3 %>%  mutate(grp=case_when(
  smoker=="nonsmoker" & expr_sex=="female" ~ "num_ns_f",
  smoker=="nonsmoker" & expr_sex=="male" ~ "num_ns_m",
  smoker=="smoker" & expr_sex=="female" ~ "num_s_f",
  smoker=="smoker" & expr_sex=="male" ~ "num_s_m")) %>% 
  group_by(grp) %>% 
  count() %>%
  ungroup() %>%
  #pivot_wider(names_from=grp, values_from=n) %>% 
  mutate(study_acc=study_name) %>% 
  select(study_acc, everything())

counts_by_grp 


# Output #1:
# append the counts to the dataset list as a sheet
# if you have problems with this, you can just write it out and paste it into the file
ss <- sheets_find("ds_list")
#sheets_write(ss,data=counts_by_grp, sheet=grp_id)  # for the first in a group
sheets_append(ss,data=counts_by_grp, sheet=grp_id) # once the sheet has been created

# Output #2:
# locally write out the labels of the individual samples with the covariates
dir.create(sprintf("data/grp_data/%s/", grp_id)) # the first time for each group

smok_w_sex_lab3 %>% left_join(covariates) %>%
  write_csv(sprintf("data/grp_data/%s/%s_metadata.csv", grp_id, study_name))

```

```
STOP HERE FOR NOW :)

### GENE LEVEL
5. Expression data QC
```{r}
```

6. Let's look at the gene x sex interactions!

6a. Smoking effects --> then gene by sex effects
```{r}
```

6b. ANOVA!! 
```{r}
```

6c.[if group data] Perform meta-analysis

```{r}

```

8.  Examine the results
```{r}

```

9. Save the results
```{r}

```