---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This is an example notebook for how to analyze an individual dataset or a group of datasets.

See the `README.md` for more information about how everything is structured.

Make a copy of this notebook and label it with:  
`ds_analysis_<ds|grp_id>.Rmd`   
It is located in:   
`code/`  
  `ds_analysis/`  

Then in your local data directory, make a directory for this group or dataset:  
`data/`  
  `group_data/`  
    `<grp_id>/`  
  `individual_data/`   
    `<ds_id>/` - in this case this the study accession


### BEFORE ANALYSIS
0. Fill in the keep/dataset comparison in `tissue_info` or `treated_cells_info`, and update the `ds_list` spreadsheet. 

For an example of this, go to the `tissue_info` spreadsheet and look at the "small airway epithelium" dataset. I added the "comparison" and "in group" columns. 
The comparison should say one of the following if we keep it ("in group" column):
* "current vs never"
* "current vs former"
* "current vs former vs never"
* "former vs never"
and then be followed with additional conditions with a ";".
Alternately if it doesn't fit in these categories, you can just describe
e.g. "current vs former; COPD" -OR- if we're not keeping it b/c there are no healthy smokers: "tumor samples". We may come back to these datasets, the goal is just to group for meta-analysis.

We then add all of the groups and their descriptions to the `ds_list` file and assign a group id. One study can belong to multiple groups.

### METADATA LEVEL
```{r}
require('GEOquery')
require('tidyverse')
require('googlesheets4') # from github! 
require('googledrive')


# you need to run `sheets_auth()` first, this will setup a browser prompt, and then you can read the sheet
# you can do this for any sheet in the google drive, you just need to have the URL
sample_sex_lab <- read_sheet("https://docs.google.com/spreadsheets/d/1lEUVsyXLDcyUQB7mXk0B-LtsWkBn66gKaIV74OFAGVs/edit#gid=463642465")
```

1. Download all the data for a dataset group
```{r}
list.ids <- c("GSE22047",  "GSE43079")
# step 1: try GEOQuery for the "GSE" prefix ones
gse.obj1 <- getGEO(list.ids[[1]]) 
gse.obj2 <- getGEO(list.ids[[2]]) 

# step 2: list ones not in available, we can grab from refine-bio
# the problem is that this is a little slow. You can also download the "E-" ones from ArrayExpress.

```



2. Identify treatment and control data
```{r}
# note: one obj may have mult platforms
# grab phenotype data
ls(gse.obj1)
p1_gpl570 <- pData(gse.obj1$`GSE22047-GPL570_series_matrix.txt.gz`)
p1_gpl6804 <- pData(gse.obj1$`GSE22047-GPL6804_series_matrix.txt.gz`)
#p2_gpl570 <- pData(gse.obj2$`GSE43079-GPL570_series_matrix.txt.gz`)
#p2_gpl16419 <- pData(gse.obj2$`GSE43079-GPL16419_series_matrix.txt.gz`)
```



### FOR EACH STUDY/PLATFORM

You will have to identify which columns are present and describe the smoker status and other covariates 
```{r}
study_name <- "GSE22047"
colnames(p1_gpl570) 
p1_gpl570 %>% View()
p1_gpl570 %>% select(geo_accession, characteristics_ch1.1, characteristics_ch1.2, characteristics_ch1.3) %>% View()

# characteristics_ch1.x often have data too -- here they are a mess tho and contain diff fields, so be careful

# always want geo_accession
# grab the columns but make things friendlier (e.g. no spaces in the names)
smok_vars <- p1_gpl570 %>% 
  select(geo_accession, "copd status:ch1", "smoking status:ch1") %>%
  rename(copd="copd status:ch1", smoker="smoking status:ch1")

# what is missing?
# what are the values?
table(smok_vars$copd)
table(is.na(smok_vars$copd)) # looks like absence means no? 

table(smok_vars$smoker)  # multiple labels mean the same thing "NS", "nonsmoker" (poor metadata), we will fix
table(is.na(smok_vars$smoker)) # all present, yay

# clean up the data
smok_vars2 <- smok_vars %>%
  mutate(smoker=case_when(
    smoker %in% c("NS", "nonsmoker") ~ "nonsmoker",
    smoker %in% c("S", "smoker") ~ "smoker"
  )) %>%
  mutate(copd=ifelse(is.na(copd), "no", copd))

# what is the breakdown?
table(smok_vars2$copd, smok_vars2$smoker)

# now for this analysis, we want health nonsmoker vs smoker
smok_healthy <- smok_vars2 %>% 
  filter(copd == "no") %>% 
  select(-copd)     

# what other covariates are present?
# -- many of these are NAs!
# -- are there sex labels somewhere? these should match "text_sex", but sometimes they do not
#summary(p1_gpl570)
covariate <- p1_gpl570 %>% 
  select(geo_accession, "serum 25-oh-d:ch1", "ethnicity:ch1") %>%
  rename(d_vit="serum 25-oh-d:ch1", eth="ethnicity:ch1")

# it's possible that there are multiple samples for the same person
# look for this!
p1_gpl570 %>%
  select(geo_accession, `department of genetic medicine id:ch1`) %>%
  rename(id=`department of genetic medicine id:ch1`) %>%
  group_by(id) %>%
  count() %>%
  arrange(desc(n))


table(covariate$d_vit)
table(is.na(covariate$d_vit)) # missing 205, this doesnt seem useful, but worth looking at to try to figure out what is going on?
# it's possible someone is present only once...
table(covariate$eth) 
table(is.na(covariate$eth)) # missing 156 samples

# write this out locally
covariate %>% write_csv("..") ##
```

3. Get the number m and number f in each category, as well as the IDs for these
```{r}
study_sex_lab <- sample_sex_lab %>% 
  filter(study_acc==study_name) %>% 
  select(-study_acc)

smok_w_sex_lab <- smok_healthy %>%
  left_join(study_sex_lab, by=c("geo_accession"="sample_acc"))

# how much is missing sex labels?
table(is.na(smok_w_sex_lab$expr_sex)) # 28 are missing, we still have a fair number
smok_w_sex_lab2 <- smok_w_sex_lab %>% filter(!is.na(expr_sex))

# get the sex breakdown
table(smok_w_sex_lab2[,c("expr_sex", "smoker")])



# if there is less than 5 in a category, make note of it on the `tissue_info` spreadsheet in a column, we may need to discard
```

```{r}

# double check the distribution of pred scores, we want the two groups to be well separated!
ggplot(smok_w_sex_lab2, aes(x=pred, y=expr_sex))+
  geom_point(aes(col=expr_sex), position=position_jitter(), alpha=0.5) + 
  xlab("P(male)")+
  ylab("")
# we do not want a ton in the middle. this one looks good


# compare to the text labels if they are present
# - switch "expr_sex" to "text_sex" to visualize

# remove any that are mislabeled
smok_w_sex_lab3 <- smok_w_sex_lab2 %>% 
  filter(text_sex==expr_sex | text_sex == "unknown")
# (here none are removed)
```

4. Write out the breakdown in two places 
```{r}
# another way to get breakdown, we will save this
counts_by_grp <- smok_w_sex_lab3 %>%  mutate(grp=case_when(
  smoker=="nonsmoker" & expr_sex=="female" ~ "num_ns_f",
  smoker=="nonsmoker" & expr_sex=="male" ~ "num_ns_m",
  smoker=="smoker" & expr_sex=="female" ~ "num_s_f",
  smoker=="smoker" & expr_sex=="male" ~ "num_s_m"
  )) %>% 
  group_by(grp) %>% 
  count() %>%
  ungroup() %>%
  pivot_wider(names_from=grp, values_from=n) %>% 
  mutate(study_acc=study_name) %>% 
  select(study_acc, everything())

counts_by_grp 


# append the counts to the dataset list as a sheet
ss <- sheets_find("ds_list")
sheets_write(ss,data=counts_by_grp, sheet="grp1_test") 
# sheets_append(ss,data=counts_by_grp, sheet="grp1_test") # once the sheet has been created

# locally write out the labels of the individual samples
grp_id <- "grp1_test"
dir.create(sprintf("data/grp_data/%s/", grp_id)) # the first time

smok_w_sex_lab3 %>% left_join(covariates) %>%
  write_csv(sprintf("data/grp_data/%s/%s_metadata.csv", grp_id, study_name))

```

```


### GENE LEVEL
5. Expression data QC
```{r}
```

6. Let's look at the gene x sex interactions!

6a. Smoking effects --> then gene by sex effects
```{r}
```

6b. ANOVA!! 
```{r}
```

6c.[if group data] Perform meta-analysis

```{r}

```

8.  Examine the results
```{r}

```

9. Save the results
```{r}

```