---
title: "06a_clean_combine_other_ds.Rmd"
author: "E Flynn"
date: "9/6/2020"
output: html_document
---

Code for cleaning the covariate data across a range of tissues.

Note: many of these studies are from the same labs and/or contain overlapping samples!


tissues:
- airway epithelium
- large airway epithelium

- lung
- bronchial epithelium
- alveolar macrophages
- bronchial alveolar lavage

- nasal epithelium
- buccal mucosa
- sputum macrophages

- blood


```{r}
library('tidyverse')
library('GEOquery')
```


```{r}
prev_data <- read_csv("data/sae_sl_mapped.csv") 
prev_data %>% separate_rows(study, sep=";") %>% distinct(study)
```

AE/large airway
GSE10006
GSE4302
GSE43939
GSE994
GSE16696
GSE11906
GSE17905
GSE18385
GSE19722


```{r}
clean_col <- function(x){
  if(!str_detect(x, ":")){
    return(x)
  }
  y <- str_split(x, ":")[[1]]
  return(str_trim(y[[2]]))
}
```

```{r}
gse <- getGEO("GSE10006") # Cornell, Strulovici-Barel




# -------- "GSE4302" ------ #

gse2 <- getGEO("GSE4302") # UCSF, Woodruff
pDat2 <- pData(gse2[[1]]) %>% select(geo_accession, source_name_ch1, characteristics_ch1, relation, `sample type:ch1`)
# reanalyzed by: GSE60486 -- double check

pDat2.2 <- pDat2 %>% 
  select(-relation, -characteristics_ch1) %>% 
  rename(smok=`sample type:ch1`) %>%
  filter(!str_detect(smok, "Asthmatic")) %>% 
  mutate(smok=case_when(smok=="Smoker" ~ "S",
                        smok=="Healthy control" ~ "NS"))

# -------- "GSE43939" ------ #

gse3 <- getGEO("GSE43939") # Cornell, Strulovici-Barel
pDat3 <- pData(gse3[[1]]) %>% 
  select(geo_accession,  source_name_ch1, `smoking status:ch1`) %>%
  rename(smok=`smoking status:ch1`) %>%
  mutate(smok=case_when(smok=="smoker"~ "S",
                        smok=="nonsmoker"~"NS"))
# NOTE: large -AND- small AE


# -------- "GSE994" ------ #

gse4 <- getGEO("GSE994") # BU, Spira

pDat4 <- pData(gse4[[1]]) %>% 
  select(geo_accession,title, relation, contains("ch1")) %>%
  select(-contains("protocol"), -contains("taxid"), -contains("organism"), -contains("molecule"))
# reanalyzed by: GSE60486 -- double check


pDat4.2 <- pDat4 %>% 
  select(-relation) %>%
  group_by(geo_accession) %>%
  mutate(across(contains("characteristics_ch1"), ~clean_col(.))) %>%
  mutate(across(everything(), ~replace_na(., replace=""))) %>%
  rename("sex"=characteristics_ch1.1, "sex2"=`sex:ch1`,
         "age"="characteristics_ch1.3", "age2"="age:ch1",
         "race"="characteristics_ch1.2", "race2"="race:ch1", 
         "pkyrs"="characteristics_ch1.4", "pkyrs2"="pkyrs:ch1",
         "history"="characteristics_ch1.5", "history2"="history:ch1",
         "patient_id"="characteristics_ch1.7", "patient_id2"="patient_id:ch1",
         "status"="characteristics_ch1.6", "status2"="status:ch1",
         "tissue"="characteristics_ch1", "tissue2"="tissue:ch1")

# all match
pDat4.2 %>%
  filter(sex!=sex2 | age != age2 | race != race2 | pkyrs != pkyrs2 |
           history != history2 | patient_id != patient_id2 | status != status2 |
           tissue != tissue2)

pDat4.2 %>% pull(source_type_ch1)
pDat4.3 <- pDat4.2 %>% 
  ungroup() %>%
  select(-contains("2"), -source_name_ch1) %>% 
  mutate(title=tolower(as.character(title))) %>% 
  mutate(diff_formatted=str_detect(title, "sample")) %>%
  mutate(title=str_replace_all(title, "sample|\\_|without cancer", " ")) %>%
  mutate(id=str_extract(title,"[0-9]+")) %>%
  mutate(smok_status=str_replace_all(title,"[0-9]+| ", "")) %>%
  mutate(smok=case_when(smok_status=="currentsmoker"~"S",
                        smok_status=="neversmoker" ~ "NS")) %>%
  select(-title, -smok_status) %>%
  mutate(sex=tolower(sex),
         race=case_when(race=="CAU" ~ "white",
                        race=="AFA"~"black",
                        race=="ASI"~"asian",
                        race=="" ~ ""))
# 9/75 have demographic info, 10 are diff formatted -- are these from a different study?
# //TODO: look at status, patient id, id, history
```


```{r}
# -------- GSE16696" ------ #

gse5 <- getGEO("GSE16696") # Cornell - Strulovici-Barel
pDat5 <-pData(gse5[[1]]) %>% 
  select( geo_accession,title, source_name_ch1,relation, 
          contains("characteristics"), "Ethnic group:ch1", "Age:ch1", "Sex:ch1", "Smoking status:ch1")
# reanalyzed by: GSE119087

pDat5.2 <- pDat5 %>% select(-relation) %>%
    group_by(geo_accession) %>%
    mutate(across(contains("characteristics_ch1"), ~clean_col(.))) %>%
    mutate(across(everything(), ~replace_na(., replace=""))) %>%
    ungroup() %>%
    rename(age="characteristics_ch1", sex="characteristics_ch1.1", race="characteristics_ch1.2",
           smok_status="characteristics_ch1.3", race2="Ethnic group:ch1", age2="Age:ch1", 
           sex2="Sex:ch1", smok_status2="Smoking status:ch1")
pDat5.3 <- pDat5.2 %>% 
  select(-contains("2"), -source_name_ch1) %>%
  mutate(title=as.character(title)) %>%
  mutate(mas5=str_detect(title, "MAS5")) %>%
  separate(title, into=c("tissue", "smok1"), sep=",") %>%
  mutate(id=str_extract(smok1, "[0-9]+"),
         smok1=str_extract(smok1, "[A-z|\\-]+")) %>%
  separate(smok_status, into=c("smok", "pkyrs"), sep=",") %>%
  mutate(pkyrs=str_extract(pkyrs, "[0-9]+"),
         sex=case_when(sex=="M" ~ "male", sex=="F"~"female"),
         race=ifelse(race=="hispnaic","hispanic", race)) 

pDat5.3 %>% filter(smok != smok1) # this is messed up. let's stick w smok
pDat5.4 <- pDat5.3 %>% select(-smok1) %>%
  mutate(smok=case_when(smok=="non-smoker"~"NS",
                        smok=="smoker"~"S"))


gse6 <- getGEO("GSE11906") # Cornell - Strulovici-Barel
# reanalyzed by: GSE60486 , GSE119087
gse7 <- getGEO("GSE17905") # Cornell - Strulovici-Barel
# reanalyzed by: GSE60486 , GSE119087
gse8 <- getGEO("GSE19722") # Cornell - Strulovici-Barel
gse9 <- getGEO("GSE18385") # Cornell - Strulovici-Barel

head(pData(gse9[[1]]))
```
SRP024274
SRP067922
SRP082973
```{r}

```


LUNG
GSE10072
GSE103174
GSE12667
GSE1650
GSE32537
GSE37768
GSE43458
GSE63882

```{r}
gse10 <- getGEO("GSE10072")
head(pData(gse10[[1]])) # NCI - Lee, adenocarcinoma study
# reanalyzed by GSE60486

pDat10 <- pData(gse10[[1]]) %>%
  select(geo_accession, title, source_name_ch1, contains("characteristics"),
         "Age at Diagnosis:ch1", "Cigarette Smoking Status:ch1", "Gender:ch1", "Stage:ch1") 

pDat10.2 <- pDat10 %>%
  group_by(geo_accession) %>%
  mutate(across(c(title, source_name_ch1),as.character)) %>%
  mutate(across(contains("characteristics_ch1"), ~clean_col(.))) %>%
  mutate(across(everything(), ~replace_na(., replace=""))) %>%
  ungroup() %>% 
  rename("sex"="characteristics_ch1", "age"="characteristics_ch1.1",
         "smok_status"="characteristics_ch1.2", "stage"="characteristics_ch1.3",
         "age2"=`Age at Diagnosis:ch1`, "smok_status2"=`Cigarette Smoking Status:ch1`, 
         "sex2"="Gender:ch1", "stage2"=`Stage:ch1`)
# all matches
pDat10.2 %>% 
  filter(sex != sex2 | age != age2 | stage != stage2 | smok_status != smok_status2) 
pDat10.3 <- pDat10.2 %>%
  select(-contains("2"), -source_name_ch1) %>%
  separate(title, into=c("cancer_status", "id"), sep="_") %>%
  mutate(sex=tolower(sex)) %>%
  mutate(smok=case_when(smok_status=="Never Smoked"~"NS", 
                        smok_status=="Current Smoker"~"S",
                        smok_status=="Former Smoker"~"FS")) %>%
  mutate(cancer=str_detect(cancer_status, "Tumor"))

# remove tumor samples???
pDat10.4 <- pDat10.3 %>% filter(!cancer, smok_status!="FS")

```

```{r}
gse11 <- getGEO("GSE103174") # IDIBAPS Barcelona - Guillaume
pData11 <- pData(gse11[[1]]) %>% 
  select(geo_accession, title, contains("ch1")) %>% 
  select(-contains("protocol"), -contains("taxid"), -contains("label"), -contains("organism"))

gse12 <- getGEO("GSE12667")
gse13 <- getGEO("GSE1650") # no demo
gse14 <- getGEO("GSE32537")
gse15 <- getGEO("GSE37768") # no demo
gse16 <- getGEO("GSE43458") # no demo
gse17 <- getGEO("GSE63882")


```

bronchial epithelium
GSE97010
GSE7895
GSE19027
```{r}
gse18 <- getGEO("GSE97010")
gse19 <- getGEO("GSE7895") # age
gse20 <- getGEO("GSE19027") # demo

```

alveolar macrophages
GSE13896
GSE13931
```{r}
gse21 <- getGEO("GSE13896") # demo
gse22 <- getGEO("GSE13931") # demo
```

bronchial alveolar lavage
GSE10038
GSE37147
GSE8823
```{r}
gse23 <- getGEO("GSE37147") # age
```



buccal mucosa
GSE17913
GSE40013
GSE8987
GSE994
```{r}

```

nasal epithelium
GSE8987
sputum macrophages
GSE46903
```{r}

```


BLOOD
GSE55962
GSE103174
GSE12585
GSE42057
GSE87072
```{r}


```

