---
title: "05_tissue_ds_covariate_table"
author: "E Flynn"
date: "7/10/2020"
output: html_document
---



```{r}
require('tidyverse')
require('googlesheets4')
require('googledrive')
require('lubridate')
require('GEOquery')
```



Read in all the covariate data
```{r}
# sheets_auth()
# my_files <- googledrive::drive_ls(path="CERSI_project/data/cleaned_covariates/gs_sheets/")
# gs_list <- my_files %>% filter(!str_detect(name, ".csv"))
# # note - I can only read these in if they are made into google sheets... SIGH
# my_sheets <- lapply(gs_list$id, function(sheet_id){
#   ss <- googledrive::drive_get(as_id(sheet_id))
#   my_df <- read_sheet(ss, sheet=1)
#   return(my_df)
# })
# names(my_sheets) <- gs_list$name

# UGH. That won't work.
# Giving up and downloading to local and the"n going from there
list_cleaned <- list.files("data/cleaned_covariates/") %>% str_replace_all( ".csv", "")
my_sheets <- lapply(list_cleaned, function(my_f){
  read_csv(sprintf("data/cleaned_covariates/%s.csv", my_f))
})
names(my_sheets) <- list_cleaned

# what is overlapping?
list_acc <- lapply(my_sheets, function(x) x$geo_accession)
length(unique(unlist(list_acc))) # 959

length(unlist(list_acc)) # 1811

# --> really need to arrange into sample, studies
# but also need the same columns to do that!
list_cols <- lapply(my_sheets, function(x) colnames(x))
unique(unlist(list_cols))
table(unlist(list_cols))

# REFORMAT
# gse108134$characteristics_ch1 --> take out copd
#   fix gse63127$copd

table(my_sheets$gse11784$copd) # n/y
table(my_sheets$gse64614$copd) # n/y
table(my_sheets$gse63127$copd) # this is NA/yes

# -- FIX values -- #
my_sheets$gse108134$copd <- str_detect(my_sheets$gse108134$characteristics_ch1, "COPD")
my_sheets$gse108134 <- my_sheets$gse108134 %>%
  mutate(copd=ifelse(copd, "y", "n")) %>%
  select(-characteristics_ch1)

my_sheets$gse63127 <- my_sheets$gse63127 %>%
  mutate(copd=ifelse(is.na(copd), NA, "y"))

# -- rename -- #
# RENAME:
# gse19667,gse63127$source_name_ch1 --> source_name
# gse108134$description --> dgm_id
# gse63127$"department of genetic medicine id:ch1" --> dgm_id
# gse108134$"30-day mean pm2.5 exposure level:ch1" --> pm2.5
# gse108134$"time of serial bronchoscopy (month):ch1" --> month

my_sheets$gse108134 <- my_sheets$gse108134 %>% 
  rename(
    dgm_id=description,
    pm2.5="30-day mean pm2.5 exposure level:ch1",
    month="time of serial bronchoscopy (month):ch1"
  )

my_sheets$gse19667 <- my_sheets$gse19667 %>% 
  rename(source_name=source_name_ch1)
my_sheets$gse63127 <- my_sheets$gse63127 %>% 
  rename(source_name=source_name_ch1, 
         dgm_id="department of genetic medicine id:ch1" )


View(my_sheets$gse64614)
gse <- GEOquery::getGEO("GSE64614") 
pDat <- pData(gse$GSE64614_series_matrix.txt.gz) 
# also has a DGM_ID we should be using

dgm_dat <- pDat %>% 
  select(geo_accession, `department of genetic medicine id:ch1`) %>%
  rename(dgm_id=`department of genetic medicine id:ch1`)
gse64614_fix <- left_join(my_sheets$gse64614, dgm_dat, by="geo_accession")
gse64614_fix2 <- gse64614_fix %>% 
  mutate(
    copd=ifelse(smok=="COPD", "y", copd),
    smok=ifelse(smok=="COPD", NA, smok),
    ethnic=case_when(ethnic=="Afr" ~ "black",
                     ethnic=="Eur" ~ "white", 
                     TRUE ~ ethnic),
    smok=case_when(smok=="nonsmoker"~ "NS",
                   smok=="smoker" ~ "S",
                   TRUE ~ smok)
    ) 

my_sheets$gse64614 <- gse64614_fix2 

gse2 <- getGEO("GSE19667")
pDat2 <- pData(gse2$GSE19667_series_matrix.txt.gz) 


gse19667_fix <- pDat2 %>% 
  unite("sex", c(`sex:ch1`, `Sex:ch1`), sep=";", na.rm=TRUE) %>%
  unite("ethnic", c("ethnic group:ch1", "Ethnic group:ch1"), sep=";", na.rm=TRUE) %>%
  unite("smok", c("smoking status:ch1", "Smoking status:ch1"), sep=";", na.rm=TRUE) %>%
  unite("age", c("age:ch1", "Age:ch1"), sep=";", na.rm=TRUE)
# add in pack years
# source_name, description, title - theyre all small airways, description is sometimes empty
# title appears to have the ID
gse19667_fix2 <- gse19667_fix %>%
  select(geo_accession, source_name_ch1, title, sex, ethnic, smok, age) %>%
  mutate(pack_years=str_extract(smok,"[0-9|\\.]+"),
         smok=case_when(str_detect(smok, "non-smoker") ~ "NS",
                        str_detect(smok, "smoker") ~ "S",
                        TRUE ~ smok),
         dgm_id=str_extract(title,"[0-9]+")) %>%
  select(-title) %>%
  rename(source_name=source_name_ch1)

my_sheets$gse19667 <- gse19667_fix2 



gse3 <- getGEO("GSE19407")
pData3 <- pData(gse3$GSE19407_series_matrix.txt.gz)
View(pData3 %>% filter(geo_accession %in% sex_lab_switch$geo_accession) )
# ok so these aren't actually switching, the processed table is wrong :/
# TODO: extract pack-years, title, COPD

 gse19407_fix <- pData3 %>% unite("sex", c(`sex:ch1`, `Sex:ch1`), sep=";", na.rm=TRUE) %>%
  unite("ethnic", c("ethnic group:ch1", "Ethnic group:ch1"), sep=";", na.rm=TRUE) %>%
  unite("smok", c("smoking status:ch1", "Smoking Status:ch1", "Smoking status:ch1"), sep=";", na.rm=TRUE) %>%
  unite("age", c("age:ch1", "Age:ch1"), sep=";", na.rm=TRUE)
# add in pack years
# source_name, description, title - theyre all small airways, description is sometimes empty
# title appears to have the ID
gse19407_fix2 <- gse19407_fix %>%
  select(geo_accession, source_name_ch1, title, sex, ethnic, smok, age) %>%
  mutate(pack_years=str_extract(smok,"[0-9|\\.]+"),
         smok=case_when(str_detect(smok, "non-smoker") ~ "NS",
                        str_detect(smok, "smoker|pack-years") ~ "S",
                        TRUE ~ smok),
         copd=ifelse(str_detect(title, "COPD"), "y", NA),
         dgm_id=str_extract(title,"[0-9]+")) %>%
  select(-title) %>%
  rename(source_name=source_name_ch1)

my_sheets$gse19407 <- gse19407_fix2 



list_cols <- lapply(my_sheets, function(x) colnames(x))
all_cols <- setdiff(names(sort(desc(table(unlist(list_cols))))), "X1")


# -- check for NAs in other columns -- #

# -- fill in missing columns -- #
add_missing_cols <- function(sheet, study_name){
  missing_cols <- setdiff(all_cols, colnames(sheet))
  for (new_col in missing_cols){
    sheet[,new_col] <- NA
  }
  sheet$study <- study_name
  return(sheet %>% select(study, all_cols))
}

reform_sheets <- lapply(1:length(my_sheets), function(i) add_missing_cols(my_sheets[[i]], names(my_sheets)[[i]]))

# -- put together -- #
sheets_comb <- do.call(rbind, reform_sheets)

collapse_id <- function(x) ifelse(all(is.na(x)), NA, paste(unique(x[!is.na(x)]), collapse=";"))

# -- condense by ID -- #
sheets_comb2 <- sheets_comb %>% 
  group_by(geo_accession) %>%
  mutate(study=collapse_id(study), # // TODO: replace with mutate_all
         smok=collapse_id(smok),
         sex=collapse_id(sex),
         source_name=collapse_id(source_name),
         ethnic=collapse_id(ethnic),
         age=collapse_id(age),
         copd=collapse_id(copd),
         dgm_id=collapse_id(dgm_id),
         vitD=collapse_id(vitD),
         cilia=collapse_id(cilia),
         month=collapse_id(month),
         pm2.5=collapse_id(pm2.5)) %>%
  unique() %>%
  ungroup()
list_leveled_cols <- c("smok", "sex", "ethnic", "source_name", "copd", "vitD", "month")
category_counts <- lapply(list_leveled_cols, function(x) table(sheets_comb2[,x]))
na_counts <- lapply(list_leveled_cols, function(x) table(is.na(sheets_comb2[,x])))

names(category_counts) <- list_leveled_cols
category_counts

names(na_counts) <- list_leveled_cols
na_counts 

sheets_comb3 <- sheets_comb2 %>% mutate(copd=ifelse(copd=="n;y", "y", copd))

# // TODO: I think should be Afr/Eur b/c descent?
# // TODO: redo some more of these just as a sanity check, fine for now
# // TODO: compare w lit

# FIX SOURCE NAME ? it's not super informative in most cases
#sheets_comb2 %>% filter(source_name=="small airway epithelium (SAE), NS") %>% select(study) # gse63127
#sheets_comb2 %>% filter(study=="gse63127") %>% select(source_name) %>% unique() # double check

# SAVE THIS!
sheets_comb3 %>% write_csv("sae_summary.csv")
```



Get date info
```{r}
# which is the "first" instance of each? is there study membership? ASSIGN
# does one study contain all others?

gse.list <- sapply(unique(sheets_comb6$first_study), toupper)
gses <- lapply(gse.list, getGEO) # only one for each
# what are the platforms on each of these
lapply(gses, function(x) x[[1]]@annotation) # they're all GPL570.. makes our life easy
# -- each sample HAS a submission date
# extract that
submission_dates <- do.call(rbind, lapply(gses, function(x) pData(x[[1]])[,c("geo_accession", "submission_date")]))

# -- each study also has a submission date
study_dates <- do.call(rbind, lapply(gses, function(x) experimentData(x[[1]])@other[c("geo_accession", "submission_date")]))%>% as.data.frame() %>% 
  rename(study=geo_accession, study_date=submission_date) %>%
  mutate(study=as.character(study),
         study_date=mdy(as.character(study_date))) %>%
  arrange(study_date)


gsm_to_date <- submission_dates %>% 
  group_by(geo_accession) %>%
  summarize(submission_date=collapse_id(submission_date))


# none have repeats! that's good
sheets_comb4 <- sheets_comb3 %>% left_join(gsm_to_date) %>%
  mutate(submission_date=mdy(submission_date))
# covariate num_total
unique(sheets_comb4$submission_date)[order(unique(sheets_comb4$submission_date))]
# there are *TWENTY-SIX* dates
# might want to group by year/month, with some window??
# when were the studies submitted?


# assign each to the earliest study date?
sheets_comb4<- sheets_comb6
sample_to_study <- sheets_comb4 %>% select(study, geo_accession, submission_date) %>% 
  separate_rows(study, sep=";") %>%
  mutate(study=toupper(study)) %>%
  left_join(study_dates) %>%
  arrange(geo_accession, study_date) %>%
  group_by(geo_accession) %>%
  mutate(idx=n():1) %>%
  top_n(1) %>%
  ungroup()

sample_to_study %>% group_by(study) %>% count()
# GSE20257, GSE19407, GSE11952 are too tiny

```



COMPARE to our sex labels
```{r}
sl_sample <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1lEUVsyXLDcyUQB7mXk0B-LtsWkBn66gKaIV74OFAGVs/edit#gid=463642465")

sl_compare <- sl_sample %>% 
  select(sample_acc, pred, expr_sex, text_sex) %>% 
  rename(geo_accession=sample_acc) %>%
  unique() %>%
  right_join(sheets_comb3 %>% select("geo_accession", "sex")) %>%
  mutate(sex=case_when(
    is.na(sex) ~ "unknown",
    sex=="M" ~ "male", 
    sex=="F" ~ "female"
  )) %>%
  rename(metadata_sex=sex) %>%
  mutate(expr_sex=ifelse(is.na(expr_sex), "unknown", expr_sex)) %>%
  select(-text_sex)
#sl_compare %>% filter(text_sex!=metadata_sex) # these are the same yay!

sl_compare %>% 
  group_by(expr_sex, metadata_sex) %>% count()
# only 4 switch labels!
sl_compare %>% filter(expr_sex=="female" & metadata_sex=="male") # these are pretty high confidence female

# > table(sl_compare$expr_sex)
# female    male unknown 
# 263     696      94 

# // TODO: compare across studies? does this help? answer label switching?
# // TODO: try to rescue the missing data! 94 with missing samples

# add a column with sex labels
sheets_comb5 <- sheets_comb4 %>% left_join(sl_compare %>% select(geo_accession, expr_sex) %>% unique()) 

```

repeated samples?!
```{r}
sheets_comb5 %>% filter(!is.na(dgm_id)) %>% group_by(dgm_id) %>% count() %>% filter(n>1) %>% arrange(desc(n)) %>% View() 
# 192 DGM_IDs are repeated

sheets_comb5 %>% group_by(dgm_id) %>% count() %>% filter(n==1 | is.na(dgm_id)) %>% ungroup() %>% summarize(tot=sum(n)) # 476  (217 are NAs)

# 668 people

condensed <- sheets_comb5 %>% filter(!is.na(dgm_id)) %>% 
  group_by(dgm_id) %>%
  mutate(sex=collapse_id(sex),
         ethnic=collapse_id(ethnic),
         smok=collapse_id(smok),
         copd=collapse_id(copd),
         expr_sex=collapse_id(expr_sex),
         source_name=collapse_id(source_name),
         submission_date=collapse_id(submission_date)) %>%
  unique()
# this is puzzling -- there is a lot of label switching

condensed_dgm <- sheets_comb5 %>% 
  filter(str_detect(dgm_id, "DGM")) %>%
  group_by(dgm_id) %>%
  mutate(sex=collapse_id(sex),
         ethnic=collapse_id(ethnic),
         smok=collapse_id(smok),
         copd=collapse_id(copd),
         expr_sex=collapse_id(expr_sex),
         source_name=collapse_id(source_name),
         submission_date=collapse_id(submission_date)) %>%
  unique()

# THIS SEEMS REASONABLE, no label switching

# // TODO: figure out what is going on with the other IDs, unclear what is going on
```


All studies, inclusive
```{r}
sheets_comb5 %>% unlist()

```

Mapped to earliest 
```{r}

```


sex breakdown table
```{r}
#  male vs female, smok vs non-smok
# by study and total
sheets_comb5 %>% filter(!is.na(smok) & expr_sex!="unknown") %>% group_by(expr_sex, smok) %>% count()
#   expr_sex smok      n
#   <chr>    <chr> <int>
# 1 female   NS      111
# 2 female   S       151
# 3 male     NS      156
# 4 male     S       540
kept_samples <- sheets_comb5 %>% filter(!is.na(smok) & expr_sex!="unknown" & copd!="y")

# exclusion: COPD, na(smok), unknown expr_sex
kept_samples %>% group_by(expr_sex, smok) %>% count()
#   expr_sex smok      n
#   <chr>    <chr> <int>
# 1 female   NS      101
# 2 female   S        90
# 3 male     NS      143
# 4 male     S       282

# // TODO: test if different - include pvals
kept_samples %>% group_by(expr_sex, smok, ethnic) %>% count()
kept_samples %>% group_by(expr_sex, smok) %>% summarize(age=mean(as.numeric(age), na.rm=TRUE))
kept_samples %>% group_by(expr_sex, smok) %>% summarize(missing_age=sum(is.na(age)))

sheets_comb6 <- sheets_comb5 %>%
  mutate_at(c("smok", "ethnic", "source_name", "sex", "expr_sex", "copd", "vitD"), ~as.factor(.)) %>%
  mutate_at(vars(age, cilia, pack_years), ~as.numeric(.)) %>%
  left_join(sample_to_study %>% select(study, geo_accession) %>% unique() %>% rename(first_study=study))

write_csv(sheets_comb6, "sae_sl_mapped.csv")
sheets_comb6 <- read_csv("sae_sl_mapped.csv") %>%
    mutate_at(c("smok", "ethnic", "source_name", "sex", "expr_sex", "copd", "vitD"), ~as.factor(.)) 
 #   smok                      ethnic      sex         expr_sex     copd    
 # NS  :299   asian               :  4   F   : 96   female :263   n   :667  
 # S   :753   black               :217   M   :229   male   :696   y   :211  
 # NA's:  1   black/hispanic;black:  1   NA's:728   unknown: 94   NA's:175  
 #            hispanic            : 50                                      
 #            white               :119                                      
 #            NA's                :662                                      
 #   vitD     
 # high:   8  
 # low :   9  
 # mid :   8  
 # NA's:1028 

 #      age          cilia         pack_years    
 # Min.   :19    Min.   :4.567   Min.   :  0.50  
 # 1st Qu.:38    1st Qu.:6.322   1st Qu.: 17.62  
 # Median :45    Median :6.788   Median : 25.00  
 # Mean   :43    Mean   :6.748   Mean   : 29.33  
 # 3rd Qu.:48    3rd Qu.:7.228   3rd Qu.: 35.75  
 # Max.   :73    Max.   :8.578   Max.   :119.00  
 # NA's   :728   NA's   :1002    NA's   :951  
#sheets_comb5 %>% write_csv("sae_summary_w_sl.csv")


```

SUMMARIZE by study
```{r}
# // TODO: what are the covariates included per study?

```

Download all of these!
```{r}
require("MetaIntegrator")
gses2 <- getGEOData(gse.list)
sapply(gses2$originalData, function(x) x$exp_comment)
# GSE108134 is missing
# do meta-analysis w/ these
sapply(gses2$originalData, function(x) dim(x$expr))

exp_mat <- do.call(cbind, lapply(gses2$originalData, function(x) x$expr))
dim(exp_mat) # only 1406 columns
save(exp_mat, file="sae_exp_mat.RData")
load("sae_exp_mat.RData")
# //TODO: what happens with repeated data?


# 1. deal w/ missing data
missing_counts <- apply(exp_mat, 2, function(x) sum(is.na(x)))

# ok - well this makes things easy - just exclude missing 
#table(missing_counts)
#missing_counts
#    0 54675 
# 1252   154 

exp_mat2 <- exp_mat[,missing_counts==0]

# remove duplicates
unique_cols <- unique(unlist(colnames(exp_mat2)))
exp_mat3 <- exp_mat2[,!duplicated(colnames(exp_mat2))]
dim(exp_mat3) # 460 columns

# there are duplicates still... BOOO
exp_mat4 <- exp_mat3[,!duplicated(t(exp_mat3))] # --> 406
save(exp_mat4, file="sae_exp_mat2.RData")

# what is the breakdown?? we had so many samples before :(
# look at what is present
# -- what wound up being duplicates??

sheets7 <- sheets_comb6 %>% 
  group_by(geo_accession) %>%
  mutate_all(~collapse_id(as.character(.))) %>%
  unique() %>%
  ungroup() %>%
  left_join(study_dates, by=c("first_study"="study")) %>%
  arrange(study_date) %>%
  mutate(study_year=year(study_date)) %>%
  mutate(year_bin=ifelse(study_year==2010, 2009, study_year)) %>%
  select(-first_study_alt, -study_year)
  
sheets8 <- sheets7 %>%
  filter(!is.na(smok)) %>%
  mutate_at(vars(smok, ethnic, sex, expr_sex, copd, month, year_bin, vitD, first_study, source_name), ~fct_explicit_na(as.factor(.))) %>%
  mutate_at(vars(age, pack_years, pm2.5, cilia), ~as.numeric(.)) %>%
  mutate(submission_date=ymd(submission_date)) %>%
  mutate(study=toupper(study)) %>%
  rename(metadata_sex=sex,
         "race/ethnicity"=ethnic)


sheets8 %>% write_csv("sae_dedup.csv")

# get the summary of these data
summary(sheets8 %>% select(-study, -geo_accession, -source_name, -dgm_id,-first_study, -year_bin, -month, -study_date) %>%
  select(smok, expr_sex, metadata_sex, copd, everything()))

sheets8_f <- sheets8 %>% filter(copd!="y" & expr_sex!="unknown")

sheets8_f %>%
  separate_rows(study, sep=";") %>%
  group_by(study) %>%
  count() %>%
  ungroup() %>%
  summarize(sum=sum(n))

# smok/sex counts by study
sheets8_f %>%
  separate_rows(study, sep=";") %>%
  group_by(study, smok, expr_sex) %>% 
  count() %>% 
  ungroup()  %>% 
  unite(grp,c("smok", "expr_sex"), sep=" - ") %>%
  pivot_wider(names_from=grp, 
              values_from = n, 
              values_fill = list(n = 0)) %>%
   left_join(study_dates) %>%
  select(study, study_date, everything()) %>%
  arrange(study_date)

# number of studies per sam
counts_per_sample <- sheets8_f %>%
  separate_rows(study, sep=";") %>%
  group_by(geo_accession) %>%
  count() %>%
  ungroup()
ggplot(counts_per_sample)+geom_histogram(aes(x=n), binwidth=1)+
  ylab("number of samples")+
  xlab("number of studies")+
  theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))


sheets8_f %>% 
  group_by(first_study, smok, expr_sex) %>% 
  count() %>% 
  ungroup()  %>% 
  unite(grp,c("smok", "expr_sex"), sep=" - ") %>%
  pivot_wider(names_from=grp, 
              values_from = n, 
              values_fill = list(n = 0)) %>%
  left_join(study_dates, by=c("first_study"="study")) %>%
  select(first_study, study_date, everything()) %>%
  arrange(study_date)

# ethnic vs category
counts_per_re <- sheets8_f %>%
  rename(race_ethnicity=`race/ethnicity`) %>%
  select(smok, expr_sex, race_ethnicity) %>%
  mutate(race_ethnicity=ifelse(str_detect(as.character(race_ethnicity), "black"), "black", as.character(race_ethnicity))) %>%
  unite(grp,c("smok", "expr_sex"), sep=" - ") %>%
  group_by(grp, race_ethnicity) %>%
  count() %>%
  ungroup() %>%
  pivot_wider(names_from=grp, 
              values_from = n, 
              values_fill = list(n = 0)) 
  
re3 <- sheets8_f %>%
  rename(race_ethnicity=`race/ethnicity`) %>%
  select(smok, expr_sex, race_ethnicity) %>%
  mutate(race_ethnicity=ifelse(str_detect(as.character(race_ethnicity), "black"), "black", as.character(race_ethnicity)),
         expr_sex=as.character(expr_sex)) %>%
  filter(!race_ethnicity %in% c("asian", "hispanic", "(Missing)"))
  

re2 <- counts_per_re %>% filter(!race_ethnicity %in% c("asian", "hispanic", "(Missing)")) %>% as.data.frame()
rownames(re2) <- re2$race_ethnicity
re2$race_ethnicity <- NULL
chisq.res <- chisq.test(re2)
round(chisq.res$expected,2)
chisq.res

chisq.race_smok <- chisq.test(table(re3$race_ethnicity, re3$smok))
chisq.race_smok
chisq.race_sex <- chisq.test(table(re3$race_ethnicity, re3$expr_sex))
chisq.race_sex
# NS

# missingness??
missing_re <- sheets8_f %>%
  rename(race_ethnicity=`race/ethnicity`) %>%
  select(smok, expr_sex, race_ethnicity) %>%
    mutate( expr_sex=as.character(expr_sex)) %>%
  mutate(race_missing=ifelse(race_ethnicity=="(Missing)", "y", "n"))
missing_re %>%
  unite(grp,c("smok", "expr_sex"), sep=" - ") %>%
  group_by(grp, race_missing) %>%
  count() %>%
  ungroup() %>%
  pivot_wider(names_from=grp, 
              values_from = n, 
              values_fill = list(n = 0)) 

chisq.race_m_smok <- chisq.test(table(missing_re$smok, missing_re$race_missing)) # p = 0.01
# more smokers with missing data than expected

chisq.race_m_sex <- chisq.test(table(missing_re$expr_sex, missing_re$race_missing)) # p = 0.8

missing_age <- sheets8_f %>%
  select(smok, expr_sex, age) %>%
    mutate( expr_sex=as.character(expr_sex)) %>%
  mutate(age_missing=ifelse(is.na(age), "y", "n"))

missing_age %>%
  unite(grp,c("smok", "expr_sex"), sep=" - ") %>%
  group_by(grp, age_missing) %>%
  count() %>%
  ungroup() %>%
  pivot_wider(names_from=grp, 
              values_from = n, 
              values_fill = list(n = 0)) 

chisq.age_m_sex <- chisq.test(table(missing_age$expr_sex, missing_age$age_missing))
chisq.age_m_sex
chisq.age_m_smok <- chisq.test(table(missing_age$smok, missing_age$age_missing))
chisq.age_m_smok # p=0.049
# more smokers with missing data than expected


# age vs category
# missing age vs category
sheets8_f %>%
  select(smok, expr_sex, age) %>%
  unite(grp,c("smok", "expr_sex"), sep=" - ") %>%
  ggplot()+
  geom_density(aes(x=age, col=factor(grp)))+
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.title = element_blank())

sheets8_f %>%
  select(smok, expr_sex, age) %>%
  unite(grp,c("smok", "expr_sex"), sep=" - ") %>%
  ggplot(aes(x=factor(grp), y=age))+
  geom_violin()+
    geom_boxplot(width=0.1)+
  geom_point(position=position_jitter(0.15), alpha=0.5)+theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"))+
  xlab("")

aov_in <- sheets8_f %>%
  select(smok, expr_sex, age) %>%
  filter(!is.na(age))

two.way <- aov(age ~ smok+expr_sex+smok*expr_sex, data = aov_in)
summary(two.way)



sheets8_f %>%
  select(smok, expr_sex, age) %>%
  unite(grp,c("smok", "expr_sex"), sep=" - ") %>%
  mutate(missing_age=(is.na(age))) %>%
  group_by(grp, missing_age) %>%
  count() %>%
  ungroup() %>%
  pivot_wider(names_from=grp, 
              values_from = n, 
              values_fill = list(n = 0)) 
# statistical test!

sheets8_f %>% 
  group_by(year_bin, smok, expr_sex) %>% 
  count() %>% 
  ungroup()  %>% 
  unite(grp,c("smok", "expr_sex"), sep=" - ") %>%
  pivot_wider(names_from=grp, 
              values_from = n, 
              values_fill = list(n = 0))

# filter
pheno_mat <- sheets8 %>%
  filter(geo_accession %in% colnames(exp_mat4)) %>%
  arrange(geo_accession) %>% 
  filter((is.na(copd) |copd=="n") & !is.na(smok)) 
summary(pheno_mat)

exp_mat5 <- exp_mat4[,pheno_mat$geo_accession]
# --> 383 samples


dmat <- dist(t(exp_mat5))
hc <- hclust(dmat)
plot(hc)
```



```{r}
# 2. visualize the data: MDS, PCA, UMAP/tSNE
require('Rtsne')

tsne_out <- Rtsne(t(as.matrix(exp_mat5)))
tsne_out2 <- cbind(data.frame(tsne_out$Y), "geo_accession"=colnames(exp_mat5)) %>%
  left_join(pheno_mat3) 

ggplot(tsne_out2, aes(x=X1, y=X2))+
  geom_point(aes(col=smok), alpha=0.7)+
  theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) # Plot the result
ggplot(tsne_out2 %>% mutate(expr_sex=factor(expr_sex, levels=c("female", "unknown", "male"))), aes(x=X1, y=X2))+
  geom_point(aes(col=expr_sex, shape=smok), alpha=0.7)+
  theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
ggplot(tsne_out2, aes(x=X1, y=X2))+
  geom_point(aes(col=first_study))
ggplot(tsne_out2 %>% mutate(year_bin=factor(year_bin)), aes(x=X1, y=X2))+
  geom_point(aes(col=year_bin), alpha=0.7)+
  theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

ggplot(tsne_out2, aes(x=X1, y=X2))+
  geom_point(aes(col=submission_date), alpha=0.7)+
  theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

# // TODO: condense dates?

# play with parameters for tsne

# add colors!!!

# PCA
save(exp_mat5, pheno_mat, file="tmp_sae_dat.RData")

require('ggfortify')
pcs <- prcomp(t(as.matrix(exp_mat5)), scale. = TRUE)
autoplot(pcs)

autoplot(pcs, data=pheno_mat, colour='smok')
autoplot(pcs, data=pheno_mat %>% mutate(expr_sex=factor(expr_sex, levels=c("female", "unknown", "male"))), colour='expr_sex', shape='smok', alpha=0.7)
autoplot(pcs, data=pheno_mat3 %>% mutate(year_bin=factor(year_bin)), colour='year_bin', alpha=0.5)
autoplot(pcs, data=pheno_mat, colour='submission_date')

# BIN SUBMISSION DATES
pheno_mat2 <- pheno_mat %>% mutate(binned_date=case_when(
submission_date %in% c("2006-03-20", "2006-06-12") ~ "grp01",
submission_date %in% c("2007-05-17") ~ "grp02",
submission_date %in% c("2007-12-21", "2008-01-03", "2008-01-10") ~ "grp03",
submission_date %in% c("2008-06-13", "2008-06-26", "2008-07-01") ~ "grp04",
submission_date %in% c("2008-12-12", "2009-01-23") ~ "grp05",
submission_date %in% c("2009-05-29", "2009-07-28") ~ "grp06", 
submission_date %in% c("2009-10-02", "2009-11-12", "2009-12-28") ~ "grp07",
submission_date %in% c("2010-06-03") ~ "grp08",
submission_date %in% c("2010-11-30") ~ "grp09",
submission_date %in% c("2011-08-17") ~ "grp10",
submission_date %in% c("2012-12-20", "2013-01-31") ~ "grp11",
submission_date %in% c("2013-11-08", "2014-11-10", "2014-12-31") ~"grp12")) 
autoplot(pcs, data=pheno_mat2, colour='binned_date')

# MDS
```



COUNTS
```{r}
pheno_mat2 %>% filter(expr_sex!="unknown") %>% group_by(binned_date, smok, expr_sex) %>% count() %>% ungroup() %>% unite(grp,c("smok", "expr_sex"), sep=" - ") %>% pivot_wider(names_from=grp, values_from = n, values_fill = list(n = 0))

study_counts <- pheno_mat2 %>%  filter(expr_sex!="unknown") %>% group_by(first_study, smok, expr_sex) %>% count() %>% ungroup() 
  

count_table_yr <- study_counts %>% unite(grp,c("smok", "expr_sex"), sep=" - ") %>% pivot_wider(names_from=grp, values_from = n, values_fill = list(n = 0)) %>%
  left_join(study_dates, by=c("first_study"="study")) %>%
  arrange(study_date) %>%
  mutate(study_year=year(study_date)) %>%
  mutate(year_bin=ifelse(study_year==2010, 2009, study_year)) %>% select(-first_study, -study_date, -study_year) %>% group_by(year_bin) %>% summarize_all(~sum(.))

study_bins <- study_dates %>% mutate(study_year=year(study_date)) %>%
  mutate(year_bin=ifelse(study_year==2010, 2009, study_year))

pheno_mat3 <- pheno_mat2 %>% left_join(study_bins %>% select(study, study_date, year_bin) %>% rename(first_study=study))


sheets_comb6 %>% 
  filter(expr_sex!="unknown" & !is.na(smok) & (is.na(copd) | copd=="n")) %>% separate_rows(study, sep=";") %>% 
  group_by(study, smok, expr_sex) %>% count() %>% ungroup() %>% unite(grp,c("smok", "expr_sex"), sep=" - ") %>% pivot_wider(names_from=grp, values_from = n, values_fill = list(n = 0))



sheets_comb6 %>% 
  filter(expr_sex!="unknown" & !is.na(smok) & (is.na(copd) | copd=="n")) %>% separate_rows(study, sep=";") %>% 
  group_by(first_study, smok, expr_sex) %>% count() %>% ungroup() %>% unite(grp,c("smok", "expr_sex"), sep=" - ") %>% pivot_wider(names_from=grp, values_from = n, values_fill = list(n = 0))


sheets_comb6$first_study_alt  <- sapply(sheets_comb6$study, function(x) str_split(x, ";")[[1]][[1]])

sheets_comb6 %>% 
  filter(expr_sex!="unknown" & !is.na(smok) & (is.na(copd) | copd=="n")) %>% separate_rows(study, sep=";") %>% 
  group_by(first_study_alt, smok, expr_sex) %>% count() %>% ungroup() %>% unite(grp,c("smok", "expr_sex"), sep=" - ") %>% pivot_wider(names_from=grp, values_from = n, values_fill = list(n = 0))

# ok... this is not right

#  Jul 13, 2009
# GSE11784
# GSE108134 
```

Setup for meta-analysis
```{r}
mapped_dat <- pheno_mat3 %>%
  mutate(year_bin=factor(paste("y",year_bin, sep=""))) %>% 
  select(geo_accession, year_bin) %>%
  arrange(year_bin, geo_accession)

exp2008 <-exp_mat5[,(mapped_dat %>% filter(year_bin=="y2008"))$geo_accession]
exp2009 <-exp_mat5[,(mapped_dat %>% filter(year_bin=="y2009"))$geo_accession]
exp2014 <-exp_mat5[,(mapped_dat %>% filter(year_bin=="y2014"))$geo_accession]

p2008 <- pheno_mat3 %>% filter(year_bin==2008) %>% arrange(geo_accession)
p2009 <- pheno_mat3 %>% filter(year_bin==2009) %>% arrange(geo_accession)
p2014 <- pheno_mat3 %>% filter(year_bin==2014) %>% arrange(geo_accession)

gse_ex <- getGEOData("GSE11784")
gse_ex2 <- gse_ex$originalData$GSE11784

gse2008 = gse_ex2
gse2008$pheno <- data.frame(p2008)
rownames(gse2008$pheno) <- p2008$geo_accession
gse2008$expr <- exp2008

gse2009 = gse_ex2
gse2009$pheno <- data.frame(p2009)
rownames(gse2009$pheno) <- p2009$geo_accession
gse2009$expr <- exp2009

gse2014 = gse_ex2
gse2014$pheno <- data.frame(p2014)
rownames(gse2014$pheno) <- p2014$geo_accession

gse2014$expr <- exp2014

# add class labels
gse2008$class <- ifelse(gse2008$pheno$smok=="S", 1, 0)
names(gse2008$class) <- gse2008$pheno$geo_accession

gse2009$class <- ifelse(gse2009$pheno$smok=="S", 1, 0)
names(gse2009$class) <- gse2009$pheno$geo_accession

gse2014$class <- ifelse(gse2014$pheno$smok=="S", 1, 0)
names(gse2014$class) <- gse2014$pheno$geo_accession

checkDataObject(gse2008, "Dataset")
checkDataObject(gse2009, "Dataset")
checkDataObject(gse2014, "Dataset")
metaObj <- list("originalData"=list("y2008"=gse2008, "y2009"=gse2009, "y2014"=gse2014))

metaObj <- runMetaAnalysis(metaObj)
metaObj <- filterGenes(metaObj)
summarizeFilterResults(metaObj, getMostRecentFilter(metaObj))

violinPlot(metaObj$filterResults$FDR0.05_es0_nStudies1_looaTRUE_hetero0, gse2008, labelColumn = 'smok')

violinPlot(metaObj$filterResults$FDR0.05_es0_nStudies1_looaTRUE_hetero0, gse2009, labelColumn = 'smok')
```


Separate by sex
```{r}




p2008_f <- pheno_mat3 %>% filter(year_bin==2008  & expr_sex=="female") %>% arrange(geo_accession)
p2009_f <- pheno_mat3 %>% filter(year_bin==2009  & expr_sex=="female") %>% arrange(geo_accession)
p2014_f <- pheno_mat3 %>% filter(year_bin==2014  & expr_sex=="female") %>% arrange(geo_accession)

exp2008_f <-exp_mat5[,p2008_f$geo_accession]
exp2009_f <-exp_mat5[,p2009_f$geo_accession]
exp2014_f <-exp_mat5[,p2014_f$geo_accession]

p2008_m <- pheno_mat3 %>% filter(year_bin==2008  & expr_sex=="male") %>% arrange(geo_accession)
p2009_m <- pheno_mat3 %>% filter(year_bin==2009  & expr_sex=="male") %>% arrange(geo_accession)
p2014_m <- pheno_mat3 %>% filter(year_bin==2014  & expr_sex=="male") %>% arrange(geo_accession)

exp2008_m <-exp_mat5[,p2008_m$geo_accession]
exp2009_m <-exp_mat5[,p2009_m$geo_accession]
exp2014_m <-exp_mat5[,p2014_m$geo_accession]

gse2008_f = gse_ex2
gse2008_f$pheno <- data.frame(p2008_f)
rownames(gse2008_f$pheno) <- p2008_f$geo_accession
gse2008_f$expr <- exp2008_f

gse2008_m = gse_ex2
gse2008_m$pheno <- data.frame(p2008_m)
rownames(gse2008_m$pheno) <- p2008_m$geo_accession
gse2008_m$expr <- exp2008_m

gse2009_f = gse_ex2
gse2009_f$pheno <- data.frame(p2009_f)
rownames(gse2009_f$pheno) <- p2009_f$geo_accession
gse2009_f$expr <- exp2009_f

gse2009_m = gse_ex2
gse2009_m$pheno <- data.frame(p2009_m)
rownames(gse2009_m$pheno) <- p2009_m$geo_accession
gse2009_m$expr <- exp2009_m

gse2014_f = gse_ex2
gse2014_f$pheno <- data.frame(p2014_f)
rownames(gse2014_f$pheno) <- p2014_f$geo_accession
gse2014_f$expr <- exp2014_f



gse2014_m = gse_ex2
gse2014_m$pheno <- data.frame(p2014_m)
rownames(gse2014_m$pheno) <- p2014_m$geo_accession
gse2014_m$expr <- exp2014_m
gse2014_m$formattedName <- "y2014"
gse2014_f$formattedName <- "y2014"

# add class labels
gse2008_f$class <- ifelse(gse2008_f$pheno$smok=="S", 1, 0)
names(gse2008_f$class) <- gse2008_f$pheno$geo_accession
gse2009_f$class <- ifelse(gse2009_f$pheno$smok=="S", 1, 0)
names(gse2009_f$class) <- gse2009_f$pheno$geo_accession
gse2014_f$class <- ifelse(gse2014_f$pheno$smok=="S", 1, 0)
names(gse2014_f$class) <- gse2014_f$pheno$geo_accession

gse2008_m$class <- ifelse(gse2008_m$pheno$smok=="S", 1, 0)
names(gse2008_m$class) <- gse2008_m$pheno$geo_accession
gse2009_m$class <- ifelse(gse2009_m$pheno$smok=="S", 1, 0)
names(gse2009_m$class) <- gse2009_m$pheno$geo_accession
gse2014_m$class <- ifelse(gse2014_m$pheno$smok=="S", 1, 0)
names(gse2014_m$class) <- gse2014_m$pheno$geo_accession


metaObj_f <- list("originalData"=list("y2008"=gse2008_f, "y2009"=gse2009_f, "y2014"=gse2014_f))
metaObj_m <- list("originalData"=list("y2008"=gse2008_m, "y2009"=gse2009_m, "y2014"=gse2014_m))
checkDataObject(metaObj_f, "Meta", "Pre-Analysis")
checkDataObject(metaObj_m, "Meta", "Pre-Analysis")

metaObj_f <- runMetaAnalysis(metaObj_f)
metaObj_f <- filterGenes(metaObj_f, effectSizeThresh = 0.5)
res_f <- summarizeFilterResults(metaObj_f, getMostRecentFilter(metaObj_f))

metaObj_m <- runMetaAnalysis(metaObj_m)
metaObj_m <- filterGenes(metaObj_m, effectSizeThresh = 0.5)
res_m <- summarizeFilterResults(metaObj_m, getMostRecentFilter(metaObj_m))

violinPlot(metaObj_f$filterResults$FDR0.05_es0.5_nStudies1_looaTRUE_hetero0, gse2014_f, labelColumn = 'smok')

violinPlot(metaObj_m$filterResults$FDR0.05_es0.5_nStudies1_looaTRUE_hetero0, gse2014_m, labelColumn = 'smok')

length(rownames(res_f$pos)) # 62
length(rownames(res_m$pos)) # 134
length(rownames(res_f$neg)) # 130
length(rownames(res_m$neg)) # 134
length(intersect(rownames(res_f$neg), rownames(res_m$neg))) # 25
length(intersect(rownames(res_f$pos), rownames(res_m$pos))) # 20

# none are flipped
intersect(rownames(res_f$pos), rownames(res_m$neg))
intersect(rownames(res_m$pos), rownames(res_f$neg))

setdiff(rownames(res_f$pos), rownames(res_m$pos))
setdiff(rownames(res_m$pos), rownames(res_f$pos))

all_sig <- unique(c(rownames(res_m$pos), rownames(res_f$pos), rownames(res_f$neg), rownames(res_m$neg)))
all_res_f <- metaObj_f$metaAnalysis$pooledResults
all_res_f$gene <- rownames(all_res_f)
all_res_m <- metaObj_m$metaAnalysis$pooledResults
all_res_m$gene <- rownames(all_res_m)

comb_res <- all_res_f %>% select(gene, effectSize, effectSizeStandardError, effectSizePval) %>% left_join(all_res_m %>% select(gene, effectSize, effectSizeStandardError, effectSizePval), by="gene")
ggplot(comb_res, aes(x=effectSize.x, y=effectSize.y))+geom_point(alpha=0.2)+geom_smooth()

ggplot(comb_res %>% filter(gene %in% all_sig), aes(x=effectSize.x, y=effectSize.y))+geom_point( alpha=0.2)+geom_errorbar(aes(ymin=effectSize.y-effectSizeStandardError.y, ymax=effectSize.y+effectSizeStandardError.y), alpha=0.3)+
  geom_errorbarh(aes(xmin=effectSize.x-effectSizeStandardError.x, xmax=effectSize.x+effectSizeStandardError.x), alpha=0.3)

all_res_f2 <- all_res_f %>%  select(gene, effectSize, effectSizeStandardError, effectSizePval) %>%  mutate(l_es=effectSize-1.96*effectSizeStandardError, u_es=effectSize+1.96*effectSizeStandardError)

all_res_m2 <- all_res_m%>%  select(gene, effectSize, effectSizeStandardError, effectSizePval) %>% mutate(l_es=effectSize-1.96*effectSizeStandardError, u_es=effectSize+1.96*effectSizeStandardError)

res3 <- all_res_f2 %>% inner_join(all_res_m2, by=c("gene"))

diff_es <- res3 %>% filter(gene %in% all_sig) %>% filter(l_es.y > u_es.x | l_es.x > u_es.y)
require('ggrepel')
ggplot(diff_es, aes(x=effectSize.x, y=effectSize.y))+
    geom_hline(yintercept = 0, linetype=2)+
    geom_vline(xintercept = 0, linetype=2)+
        geom_hline(yintercept = 0.5, linetype=3, col="red")+
    geom_vline(xintercept = 0.5, linetype=3, col="red")+
        geom_hline(yintercept = -0.5, linetype=3, col="red")+
    geom_vline(xintercept = -0.5, linetype=3, col="red")+
  geom_point(alpha=0.7, color="blue")+
  geom_errorbar(aes(ymin=l_es.y, ymax=u_es.y), alpha=0.2)+
  geom_errorbarh(aes(xmin=l_es.x, xmax=u_es.x),alpha=0.2)+
  geom_label_repel(aes(label=gene), size=3)+
  ylim(-2, 2)+
  xlim(-2, 2)+
  ylab("effect size (S-NS) - males")+
  xlab("effect size (S-NS) - females")+
  theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))


ggplot(res3 %>% filter(gene %in% all_sig), aes(x=effectSize.x, y=effectSize.y))+
    geom_hline(yintercept = 0, linetype=2)+
    geom_vline(xintercept = 0, linetype=2)+
      geom_hline(yintercept = 0.5, linetype=3, col="red")+
    geom_vline(xintercept = 0.5, linetype=3, col="red")+
        geom_hline(yintercept = -0.5, linetype=3, col="red")+
    geom_vline(xintercept = -0.5, linetype=3, col="red")+
    geom_abline(intercept=0, slope=1)+
  geom_point(alpha=0.7)+
  geom_errorbar(aes(ymin=l_es.y, ymax=u_es.y), alpha=0.2)+
  geom_errorbarh(aes(xmin=l_es.x, xmax=u_es.x),alpha=0.2)+
  ylim(-2, 2)+
  xlim(-2, 2)+
  ylab("effect size (S-NS) - males")+
  xlab("effect size (S-NS) - females")+
  theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))



diff_es2 <- diff_es %>% arrange(desc(abs(effectSize.x-effectSize.y))) 

plot.new()
metaObj_f$originalData$y2008$formattedName <- "y2008"
metaObj_f$originalData$y2009$formattedName <- "y2009"
metaObj_f$originalData$y2014$formattedName <- "y2014"
metaObj_m$originalData$y2008$formattedName <- "y2008"
metaObj_m$originalData$y2009$formattedName <- "y2009"
metaObj_m$originalData$y2014$formattedName <- "y2014"
plot.new()
forestPlot(metaObj_m, "BTNL8" )
plot.new()
forestPlot(metaObj_f, "BTNL8" )

plot.new()
forestPlot(metaObj_m, "CYP2A6" )
forestPlot(metaObj_f, "CYP2A6" )
gene_name <- "ZNF658"

es_f <- metaObj_f$metaAnalysis$datasetEffectSizes[gene_name,]
se_f <- metaObj_f$metaAnalysis$datasetEffectSizeStandardErrors[gene_name,]

pooled_df <- bind_rows(all_res_f[gene_name, c("effectSize", "effectSizeStandardError")],
          all_res_m[gene_name, c("effectSize", "effectSizeStandardError")])

colnames(pooled_df) <- c("es", "se")
pooled_df$sex <- c("female", "male")
pooled_df$ds <- "summary"

es_m <- metaObj_m$metaAnalysis$datasetEffectSizes[gene_name,]
se_m <- metaObj_m$metaAnalysis$datasetEffectSizeStandardErrors[gene_name,]

df_f <- data.frame(cbind("es"=es_f, "se"=se_f) )
df_f$ds <- rownames(df_f)
df_f$sex <- "female"
df_m <- data.frame(cbind("es"=es_m, "se"=se_m) )
df_m$ds <- rownames(df_m)
df_m$sex <- "male"
# add the pooled


df2 <- df_f %>% bind_rows(df_m) %>% bind_rows(pooled_df) %>% mutate(ds=factor(ds, levels=c("y2008", "y2009", "y2014", "summary")))

ggplot(df2, aes(y=sex, x=es, col=sex))+
  geom_point()+
  geom_vline(xintercept=0, linetype=2)+
  geom_errorbarh(aes(xmin=es-1.96*se, xmax=es+1.96*se), height=0.1)+theme_bw() +
    facet_grid(ds ~., switch="y") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(angle = 0),
    axis.line = element_line(colour = "black"))+
  ggtitle(gene_name)+
  ylab("")+
  xlab("effect size (S-NS)")
```


Read in the new one and look at:
```{r}

```


```{r}
# 3. hierachical clustering/heatmap


```

Co-normalize all of it?
```{r}

```


READ IN MISSING GSE...
```{r}

```





Start downloading the other data
```{r}

```




Look at airway epithelium, large + double check
```{r}

```


Do we have an updated study case/ctl/sex breakdown file?
If not - make this

```{r}

```